My wonderful paper

### 2024-12-09 index.qmd
Discussion

While some checks may be crucial and directly indicate an unexpected outcome, others may be tangential to the problem at hand and not indicate a root cause of an unexpected outcome. In this section, we propose a procedure to measure the effectiveness of checks that, when combined using logical operators, contribute to an unexpected outcome. A small set of independent checks is considered effective if it translates to unexpected outcomes and @fig-metric-calc provides an overview of the calculation.

@@

Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study]. Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference]. The example demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the checks to diagnose an unexpected PM10 coefficient from the generalized linear model.

This PM10 coefficient expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005] and 0 if it is within. Multiple factors can affect the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature. Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate. Outliers in the three variables can also leverage the coefficient. While these are possible factors that could affect the analysis result, it is not clear the cutoff values for these checks to determine a failure. Here we create a list of checks in @tbl-checks with varied cutoff values for each:

@@

To generate replicate of the data, we first simulate the correlation matrix of the three variables in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size. The multivariate normal distribution is transformed using the normal CDF before the inverse CDF of the assumed distributions of the three variables is applied. To determine the appropriate distribution of each variable, various distributions are fitted and compared. This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$. AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit. AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature. To include the effect of outlier, we add a single outlier to the data for both variable mortality and PM10.

@@

@@

```{r}
#| fig.width: 8
#| fig.heigh: 3
fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
                                seed = 6, nleaves = 4)
plot(fit) 
```

@@

A logic regression is fitted to predict whether the PM10 coefficient is unexpected using the checks listed in @tbl-checks and @fig-linear-reg-tree shows the logic regression tree from the fitted model. Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in  @tbl-linear-reg. The logic regression model picks up the following cutoff value for each type of check:

-   sample size *larger than* 200
-   mortality-temperature correlation greater than -0.35
-   mortality-PM10 correlation less than -0.04
-   mortality contain outliers that are detected by the univariate outlier detection

<!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->

The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature. This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient. The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.

@@

>>>>

While some checks may be crucial and directly indicate an unexpected outcome, others may be tangential to the problem at hand and not indicate a root cause of an unexpected outcome. In this section, we propose a procedure to measure the effectiveness of checks that, when combined using logical operators, contribute to an unexpected outcome. A small set of independent checks is considered effective if it translates to unexpected outcomes.

The approach relies on the use of simulated datasets that are generated based on the analyst's knowledge and assumptions about the data generation mechanism. Datasets are simulated to have the same structure and characteristics of the observed data that will be analyzed at some point in the future. For each simulated dataset, we can apply a collection of analysis validation checks to the dataset and record which ones were TRUE and which were FALSE. We can also compute the outcome of the analysis to see whether the outcome is unexpected in a given simulated datsets. We can then generate many datasets, each time applying the analysis validation checks and computing the outcome. After generating many datasets, we can relate the patterns in the analysis validation checks to the likely that the outcome will be unexpected in a given dataset. @fig-metric-calc provides an overview of the process.

@@

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.

The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10. Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between [0, 0.005], after adjusting for daily temperature [@samet200fine; @welty2005acute]. Observing an estimated coefficient outside of this interval would be highly unusual and would warrant a serious re-examination of the analysis process. Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval [0, 0.005]. In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.

This PM10 coefficient expectation can be framed as an analysis check that fails, labelled as 1, if the estimate of the PM10 coefficient (adjusted for temperature) is outside the range [0, 0.005], and 0 if it is within. Multiple factors can affect the estimated PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature. Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate. Outliers in the three variables can also leverage the coefficient. While these are possible factors that could affect the analysis result, it is not clear what the cutoff values for these checks should be to determine a failure. Here we consider a list of checks in @tbl-checks with varied cutoff values for each:

@@

## Data Simulation

To generate replicates of the dataset, we first generate the correlation matrix of the three variables (PM10, mortality, and temperature) in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size. The multivariate normal distribution is transformed using the normal CDF before the inverse CDF of the assumed distributions of the three variables is applied. To determine the appropriate distribution of each variable, various distributions are fitted and compared. This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for PM10 and temperature; and beta for PM10 after rescaling the data to $[0,1]$.

To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed dataset to refine our simulations. AIC is used to determine the best distribution fit for each variable with the QQ-plot presented in @fig-dist-fit to evaluate the fit. AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to recover the original scale), and a Weibull distribution for temperature. To include the potential effect of outliers, we add a single outlier to the data for both the mortality and PM10 variables.

@@

A logic regression is fitted using all variations of the tests in @tbl-checks to predict whether the PM10 coefficient is unexpected. @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model. Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in @tbl-linear-reg.

@@

```{r}
#| fig.width: 8
#| fig.height: 4
fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
                                seed = 6, nleaves = 4)
plot(fit) 
```

@@

As indicated in @fig-linear-reg-tree, the logic regression model picks up the following cutoff values for each type of check:

-   sample size *larger than* 200
-   mortality-temperature correlation greater than $-0.35$
-   mortality-PM10 correlation less than $-0.04$
-   mortality data contains outliers that are detected by the univariate outlier detection

<!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->
The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature. This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient. The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.

As shown in @fig-linear-reg-tree, there is no single analysis check in the tree that predicts an unexpected outcome. rather at least three checks in the tree must be TRUE in order for the model to predict an unexpected outcome. Given the high independence of the tests (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.

@@

In this paper we have developed an approach to using analysis validation checks to externalize the assumptions about the data and analysis tools made during the data analysis process. These checks can serve as a useful summary of the analyst's thought process and can desscribe how characterstics of the data may lead to unexpected outcomes. Using logic regression, we can develop a graphical summary of the analysis validation checks as well as use the logic regression fitting process to choose the optimal set of checks. The logic regression model can also be used to develop summaries of the precision and recall of the collection of analysis validation checks in predicting the likelihood of an unexpected outcome. We demonstrated our method on an example relating daily mortality to outdoor air pollution data.

An interesting connection can be drawn between our logic regression trees and a tool used in systems engineering known as a fault tree.
A fault tree is used for conducting a structured risk assessment and has a long history in aviation, aerospace, and nuclear power applications [@vesely1981fault]. A fault tree is a graphical tool that describes the possible combinations of causes and effects that lead to an anomaly. At the top of the tree is a description of an anomaly. The subsequent branches of the tree below the top event indicate possible causes of the event immediately above it in the tree. The tree can then be built recursively until we reach a root cause that cannot be further investigated. Each level of the tree is connected together using logic gates such as AND and OR gates. The leaf nodes of the tree indicate the root causes that may lead to an anomaly. While the logic regression tress are not identical to fault trees, they share many properties, such as the tree-based structure and the indicator of root causes at the leaf nodes. Perhaps more critically, both serve as graphical summaries of the assumptions made in a problem and the specific violations of those assumptions that could lead to an unexpected result. While fault trees are often used to discover the root cause of an anomaly after it occurs, an important use case for fault trees is to develop a comprehensive understanding of a system *before* an anomaly occurs [@michael2002fault].
